---
title: "plot results for publication"
author: "John Fogg & Cecile Ane"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, comment=NA)
library(ggplot2)
library(dplyr)
options(show.signif.stars=F)
```

# read data and sanity checks

read the "final" results from the simulations,
as created by step 3 `3_summarize_findings.R`:

```{r}
resultfile = "results_20221103_withD0.csv"
dat = read.csv(resultfile) %>% mutate(d=d_0) %>%
  mutate(d_0 = round(d*lambda,10),
         mu_lambda = round(mu/lambda,10), # 0.1 or 0.9
         nu_lambda = round(nu/lambda,10), # 0.5 always
         A_q = A/q,   # proportion of candidate possibly anomalous 4-taxon sets
         C_B = C/B,   # proportion of anomalous out of candidate & analyzable 4-taxon sets
         C_q = C/q,   # proportion anomalous / total number of 4-taxon sets
         A1_q = A1/q, # proportion of 4-taxon networks with a 3_2 blob
         A2_q = A2/q, # proportion of 4-taxon networks displaying 2 splits
         C1_A1 = C1/A1, # proportion anomalous, out of candidates (3_2 blob) within class 1
         C2_A2 = C2/A2)
dat$time = read.csv("results_20221103_withtime.csv")$time
```

meaning of columns:

- q: total number of 4-taxon sets
- q10, q11, q12: number of 4-taxon sets of class 1 with 0 3-blobs, 1 3-blob, 1 3_2 blob.
- q2: number of 4-taxon sets of class 2
- q3: number of 4-taxon sets of class 3
- A1=q12, A2=q2: number of quarnets susceptible of being anomalous
- A=A1+A2
- B: number of quarnets for which gene tree simulations could run
- C1, C2, C=C1+C2: number of anomalous quartets of class 1 or 2, where
  "anomalous" is simply using the empirical gene trees frequencies (no test)

sanity checks:

```{r}
stopifnot(nrow(dat) == 4*2*1*2*3*3)
# 144: number of values for: lambda, mu, nu, pi, d0, ntaxa
stopifnot(with(dat, q10+q11+q12+q2+q3 - q) == 0)
dat %>% mutate(q_expected = nsim * choose(ntaxa,4)) %>%
  filter(q != q_expected) # job 143, 6 taxa, lambda=3, d_0=1.2. time was not an issue. then what?
# stopifnot(with(dat, q == nsim * choose(ntaxa,4))) # fails: because of job 143
stopifnot(with(dat, unique(A1-q12))==0)
stopifnot(with(dat, unique(A2-q2)) ==0)
# all q quartets were analyzable:
stopifnot(with(dat, unique(B-A))==0) # to check that B=A: all quartets were analyzable
stopifnot( # check that the sum of all Ds are = B value
  dat %>% mutate(Dsum = rowSums(across(starts_with("D",ignore.case=F)))) %>%
  mutate(check = B-Dsum) %>% summarize(isDsumcorret = unique(check)) == 0
)
stopifnot(with(dat, sum((D1anom-C1)>0) == 0)) # check that D1anom <= C1
stopifnot(with(dat, sum((D2anom-C2)>0) == 0))
stopifnot(with(dat, sum((A1-C1 - D1good)<0) == 0)) # check that A1-C1 >= D1good
stopifnot(with(dat, sum((A2-C2 - D2good)<0) == 0))

dat = mutate(dat, ntaxa=factor(ntaxa), M = factor(M))
```

# averaged summary statistics

```{r}
qtot = 68800 # from this below
200*(1 + 15 + 70)/3 * (144/4/3)
datsum = dat %>% group_by(lambda,d_0) %>%
  summarize(class1 = sum(q10+q11+q12),
            class1_32blob = sum(A1),
            class2 = sum(A2),
            class3 = sum(q3),
            anomalous1_naive = sum(C1),
            anomalous2_naive = sum(C2),
            anomalous1_lib  = sum(D1anom),
            anomalous2_lib  = sum(D2anom),
            anomalous1_cons = sum(D1anom) + sum(D1ambi),
            anomalous2_cons = sum(D2anom) + sum(D2ambi),
            .groups="drop") %>%
  mutate(prop_anomalous_naive = (anomalous1_naive + anomalous2_naive)/qtot,
         prop_anomalous_lib  = (anomalous1_lib  + anomalous2_lib )/qtot,
         prop_anomalous_cons = (anomalous1_cons + anomalous2_cons)/qtot,
         prop_32     = class1_32blob/qtot,
         prop_class2 = class2/qtot,
         prob_anomalous1_naive = anomalous1_naive/class1_32blob,
         prob_anomalous2_naive = anomalous2_naive/class2,
         prob_anomalous1_lib  = anomalous1_lib /class1_32blob,
         prob_anomalous2_lib  = anomalous2_lib /class2,
         prob_anomalous1_cons = anomalous1_cons/class1_32blob,
         prob_anomalous2_cons = anomalous2_cons/class2
         )
# stopifnot(with(datsum, class1 + class2 + class3) == qtot) # last one is 68575 instead of 68800: missing 225
datsum %>% select(lambda, d_0, starts_with("prop_")) %>% arrange(prop_anomalous_naive)
ggplot(datsum, aes(x=prop_anomalous_naive, y=prop_anomalous_lib)) +
  geom_point(color="darkkhaki") +
  geom_point(aes(y=prop_anomalous_cons), color="deepskyblue") +
  geom_point(aes(y=prop_anomalous_naive), color="orangered", alpha=0.5) +
  scale_y_continuous(name="proportion anomalous", trans="sqrt") +
  scale_x_continuous(name="proportion anomalous, naive method", trans="sqrt")
  # geom_abline(intercept=0, slope=1, color="deepskyblue")
```

# plots: class proportions

proportion of 4-taxon networks with 3_2 blobs:

This visualization is not my preferred one, as we don't see the effect
of extinction. It focuses on the effect of d0, but that one is easy to see.

```{r, warning=F, eval=F}
ggplot(dat, aes(y=A1_q, x=d_0, color=factor(lambda), shape=ntaxa)) +
  geom_point(position=position_jitterdodge(dodge.width=0.2)) +
  facet_grid(~ mu_lambda) +
  theme_minimal() +
  scale_x_continuous(name="distance threshold d/\u03BB for reticulation success",
                     breaks=unique(dat$d_0), minor_breaks=NULL) +
  scale_y_continuous(name="proportion of 4-taxon subnetworks with a 3_2 blob",
                     trans="identity") # "sqrt" is too strong
```

alternative: to better see the effect of extinction.

```{r}
ggplot(dat, aes(y=A1_q, x=factor(mu_lambda), color=factor(lambda), shape=ntaxa)) +
  geom_point(position=position_jitterdodge(dodge.width=0.4), alpha=0.8) +
  guides(color=guide_legend(title=expression(speciation~lambda), order=1),
         shape=guide_legend(title="# of taxa")) +
  facet_grid(~ d_0) +
  theme_bw() + theme(plot.subtitle = element_text(hjust=0.5)) +
  labs(x=expression(turnover~rate~mu/lambda),
       subtitle=expression(distance~threshold~d/lambda),
       title=expression(Proportion~of~4*"-"*taxon~networks~with~a~3[2]~blob)) +
  scale_y_continuous(name="proportion",
                     trans="identity") # "sqrt" is too strong
ggsave("fig_prob_32blob.pdf", height=4.5, width=7)
```

proportion of 4-taxon networks with exactly 2 splits:

```{r, warning=F}
ggplot(dat, aes(y=A2_q, x=factor(mu_lambda), color=factor(lambda), shape=ntaxa)) +
  geom_point(position=position_jitterdodge(dodge.width=0.4), alpha=0.8) +
  guides(color=guide_legend(title=expression(speciation~lambda), order=1),
         shape=guide_legend(title="# of taxa")) +
  facet_grid(~ d_0) +
  theme_bw() + theme(plot.subtitle = element_text(hjust=0.5)) +
  labs(x=expression(turnover~rate~mu/lambda),
       subtitle=expression(distance~threshold~d/lambda),
       title="Proportion of 4-taxon networks of class 2") +
  scale_y_continuous(name="proportion", trans="identity") # sqrt too strong
ggsave("fig_prob_class2.pdf", height=4.5, width=7)
```

# plots: anomalies, the naive way

proportion of anomalous networks among those with a 3_2 blob:

```{r}
ggplot(filter(dat, A1>0),
       aes(y=C1_A1, x=factor(d_0), color=factor(mu_lambda), shape=ntaxa)) +
  geom_point(position=position_jitterdodge(dodge.width=0.4)) +
  # aes(alpha=cut(A1, breaks=c(0,10,200,10000)))
  guides(color=guide_legend(title=expression(turnover~mu/lambda), order=1),
         shape=guide_legend(title="# of taxa")) +
  facet_grid(~ lambda) +
  theme_bw() + theme(plot.subtitle = element_text(hjust=0.5)) +
  labs(x=expression(distance~threshold~d/lambda),
       subtitle=expression(speciation~rate~lambda),
       title=expression(Anomalous~CFs~among~class*"-"*1~networks~with~a~3[2]~blob)) +
  scale_y_continuous(name="proportion anomalous", trans="sqrt",
                     breaks=c(0,.01,.05,.1,.25,.5,.75,1), minor_breaks=c(0), expand=c(.01,.01))
ggsave("fig_prob_anomalous1_32blob.pdf", height=4.5, width=7)
```

proportion of anomalous networks among those of class 1
(including trees and networks with 2-blobs or 3_1 blobs):

```{r}
ggplot(filter(dat, q10+q11+q12 > 0),
       aes(y=C1/(q10+q11+q12), x=factor(d_0), color=factor(mu_lambda), shape=ntaxa)) +
  geom_point(position=position_jitterdodge(dodge.width=0.4)) +
  # aes(alpha=cut(A1, breaks=c(0,10,200,10000)))
  guides(color=guide_legend(title=expression(turnover~mu/lambda), order=1),
         shape=guide_legend(title="# of taxa")) +
  facet_grid(~ lambda) +
  theme_bw() + theme(plot.subtitle = element_text(hjust=0.5)) +
  labs(x=expression(distance~threshold~d/lambda),
       subtitle=expression(speciation~rate~lambda),
       title="Anomalous CFs among class-1 quarnets") +
  scale_y_continuous(name="proportion anomalous", trans="sqrt",
                     breaks=c(0,.001,.005,.01,.015,.02), minor_breaks=c(0),
                     expand=c(.002,.002))
ggsave("fig_prob_anomalous1.pdf", height=4.5, width=7)
```

proportion of anomalous networks among class-2 networks:

```{r}
ggplot(filter(dat, A2>0),
       aes(y=C2_A2, x=factor(d_0), color=factor(mu_lambda), shape=ntaxa)) +
  geom_point(position=position_jitterdodge(dodge.width=0.4)) +
  # aes(alpha=cut(C1, breaks=c(0,10,200)))
  guides(color=guide_legend(title=expression(turnover~mu/lambda), order=1),
         shape=guide_legend(title="# of taxa")) +
  facet_grid(~ lambda) +
  theme_bw() + theme(plot.subtitle = element_text(hjust=0.5)) +
  labs(x=expression(distance~threshold~d/lambda),
       subtitle=expression(speciation~rate~lambda),
       title="Anomalous CFs among class-2 quarnets") +
  scale_y_continuous(name="proportion anomalous", trans="sqrt",
                     breaks=c(0,.01,.05,.1,.25,.5,.75,1), minor_breaks=c(0),
                     expand=c(0.01,0.01))
ggsave("fig_prob_anomalous2.pdf", height=4.5, width=7)
```

## overall proportion of anomalous networks

```{r}
plot_overall_anomalies = function(dat_y){ # uses 'y' as response variable name
  ggplot(dat_y,
         aes(y, x=factor(d_0), color=factor(mu_lambda), shape=ntaxa)) +
  geom_point(position=position_jitterdodge(dodge.width=0.4)) +
  guides(color=guide_legend(title=expression(turnover~mu/lambda), order=1),
         shape=guide_legend(title="# of taxa")) +
  facet_grid(~ lambda) +
  theme_bw() + theme(plot.subtitle = element_text(hjust=0.5)) +
  labs(x=expression(distance~threshold~d/lambda),
       subtitle=expression(speciation~rate~lambda),
       title="Anomalous CFs among all 4-taxon networks") +
  scale_y_continuous(name="proportion anomalous", trans="sqrt",
                     breaks=c(0,.001,.01,.025,.05,.075,.1,.125,.25), minor_breaks=c(0),
                     expand=c(0.002,0.01))
}
```

the naive way of deciding about anomalies:

```{r}
plot_overall_anomalies(mutate(dat, y = (C1+C2)/q))
ggsave("fig_prob_anomalous12_naive.pdf", height=4.5, width=7)
```

the liberal way:

```{r}
plot_overall_anomalies(mutate(dat, y = (D1anom+D2anom)/q))
ggsave("fig_prob_anomalous12_lib.pdf", height=4.5, width=7)
```

the conservative way:

```{r}
plot_overall_anomalies(mutate(dat, y = (D1anom+D1ambi + D2anom+D2ambi)/q))
ggsave("fig_prob_anomalous12_cons.pdf", height=4.5, width=7)
```
